---
title: "Prior Predictive Distribution Sensitivity Test"
author: "Joe Brown"
date: "2023-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prior Predictive Distirbution Sensitivity Test

PPD - the distribution of data we think we would obtain before observing any data. 

```{r, message=FALSE}
library(rjags)
library(coda)
library(ggplot2)
```

## `dlnorm` on ECS

### Weak prior: writing model for `rjags`

Write model string using `dlnorm()` to estimate ECS mean and sigma (using informed prior on tau):
```{r}
mod_1 <-" model{
  
  ECS ~ dlnorm(mu, tau)
  sigma = sqrt(1 / tau)
  
  mu = 3.2
  tau ~ dgamma(4, 1)

} "
```

Set-up, initiate, and run:
```{r}
# parameters of interest
params = c("tau", "sigma", "ECS")

# initial parameter values (random value)
inits = function() { 
  inits = list("tau" = runif(1), 
               "ECS" = runif(1)) 
} 

# Compile the model - jags.model()
mod_1 = jags.model(textConnection(mod_1), data = NULL, inits = inits, n.chains = 2)

# run the mcmc sampler 
# update = burn-ins
update(mod_1, 500)

# simulate samples for parameters
mod_1_sim = coda.samples(mod_1, variable.names = params, n.iter = 15000)
```

Visualize and summarize:
```{r}
plot(mod_1_sim)
summary(mod_1_sim)
```

Sample and plot?

I think `dlnorm()` gives a value on original scale, need to take log of ECS:
```{r, eval=FALSE}
# sample the PPD
mod_1_mcmc = data.frame(
  value = generate_lognormal_samples(mu = log(29.5), sigma = 0.56, 10000), #use values from MCMC
  meth = "mcmc")

# plot samples
ggplot() +
  geom_density(data = mod_1_mcmc, aes(x = value),
               color = "dodgerblue") +
  theme_light()

# compute percentiles - compare to Sherwood table 10
percentiles = c(0.05, 0.17, 0.50, 0.83, 0.95)
result = quantile(mod_1_mcmc$value, probs = percentiles)
result
```
