---
title: "Preliminary Run Using UL and US ECS Distributions"
author: "Joe Brown"
date: "2023-09-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose and Goals 

This document is meant to serve as a preliminary analysis for the full targeteed analysis of assessing the sensitivity of climate projections to ECS distributions derived from different lines of evidence.

**Goals**:

1. Conduct MCMC sampling to estimate ECS $\mu$ and $\sigma$ parameters, using data simulated from Sherwood et al. 2020. For the preliminary run, focus on UL and US ECS distirbutions.

2. Use the parameters from the MCMC sampling procedure to simulate parameter sets for a Matilda run. 

3. Run Matilda with parameter sets (one with UL ECS distribution and one with US distribution). Parallelize this to run each 2000 times for each. Determine the time savings as well.  

4. Some plots to make - comparison of ECS distributions, Plot temperature projections with CIs (or percentiles), and compute probabilities and plot stacked bars showing the probability of warming.

Libraries:

```{r, message=FALSE}
library("rjags")
library("coda")
library("ggplot2")
library("matilda")
```

Functions:
```{r}
# estimating sigma from percentile values
estimate_sigma <- function(per_5_95, per_17_83) {
  
  # Calculate the length of the percentiles
  percentile_length_5_95 <- per_5_95[2] - per_5_95[1]
  percentile_length_17_83 <- per_17_83[2] - per_17_83[1]
  
  # Estimate sigma using percentile lengths
  sigma_5_95 <- percentile_length_5_95 / (2 * qnorm(0.95)) 
  sigma_17_83 <- percentile_length_17_83 / (2 * qnorm(0.83)) 
  
  # Average the two estimates assuming they are equally important for the estimated
  # sigma.
  estimated_sigma <- (sigma_5_95 + sigma_17_83) / 2
  
  return(estimated_sigma)
}

# reparameterizing rlnorm() parameters
lognorm <- function(m, sd){
  
  # re-parameterization of supplied mean value
  mn <- log(m^2 / sqrt(sd^2 + m^2))
  
  # re-parameterization of supplied sd value
  stdev <- sqrt(log(1 + (sd^2 / m^2)))
  
  # stores new value in list - when pushed to rlnorm(), will provide normal distribution
  # of arithmetic mean (m) and standard deviation (sd)
  c(mn, stdev)
  
}

# generates lognormal distirbutions
generate_lognormal_samples <- function(mu, sigma, n_samples) {
  
  # Generate normally distributed samples in log-space
  samples <- rlnorm(n_samples, lognorm(mu, sigma) [1], lognorm(mu, sigma) [2])
  
  return(samples)
}
```


## MCMC Sampling

In the future I want experiment automating this with a loop. Will look into using `gsub()` for this.

For now I will do them separately.

### MCMC sample for UL distribution (mu = 3.2, 5-95% = 2.3-4.7, 17-83% = 2.6-3.9)

1. Write out the heirarchical model string:

```{r}
UL_mod_string <- " model{

  for (i in 1:n) {
    y[i] ~ dnorm(mu, tau)
  }

  mu ~ dnorm(3.2, 1)
  tau ~ dgamma(5, 1 / rate)
  rate = 1.0
  sigma = sqrt(1 / tau)
} "
```

2. Simulate data and create `JAGS` data list:

```{r}
# estimate sigma
UL_sigma <- estimate_sigma(per_5_95 = c(2.3, 4.7), per_17_83 = c(2.6, 3.9))

# generate data
set.seed(123)

# simulate data
UL_dat <- generate_lognormal_samples(3.2, UL_sigma, 1000)
n <- length(UL_dat)

# JAGS data list
UL_data_jags <- list(y = UL_dat, n = n)
```

3. Specify, store parameters and create list of initial values:

```{r}
# parameter list
UL_params <- c ("mu", "tau", "sigma")

# initial values 
inits <-  function() {
  inits = list("mu" = runif(1), "tau" = runif(1), "sigma" = runif(1))
}
```

4. Intialize the model using `jags.model()`:

